SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher

### 2018-12-05 back.tex
BG and Related

The idea of prefetching begins with Jouppi's \textit{instruction stream
buffers}. Early prefetchers detected stride access patterns in order to
predict future memory requests~\cite{Smith,Baer,Stride}.  Modern prefetching
mechanisms are more sophisticated as they look into past memory
behavior~\cite{Address_Correlated,AMPM},
locality~\cite{Spatial_Pattern,SMS,Temporal_Instruction_Fetch,Off_Chip,STMS,SMS_JILP},
control-flow speculation~\cite{BFetch,MTBFetch}, and many other aspects to
detect complex access patterns in modern applications.

@@

An optimal prefetcher should be able to detect a wide range of memory access
patterns. Simple stride prefetching techniques only detect sequences of
addresses that differ by a constant value and fail to capture diverse delta
patterns. Lookahead prefetchers learn patterns by collecting histories of
observed data access patterns, and correlating these with the strides, by
considering each new page in a vacuum. Kim {\em et. al.} proposed
SPP~\cite{SPP}, a confidence-based lookahead prefetcher. SPP creates a
signature associated to a memory request by compressing the history of
accesses. By correlating the signature with future likely delta patterns, SPP
learn both simple and complicated memory access patterns quickly. Our proposed
technique builds upon SPP. Thus, we describe SPP in more detail.

>>>>

The idea of prefetching begins with Jouppi's \textit{Instruction Stream
Buffers}~\cite{ISB}. Early prefetchers detected stride access patterns in order to
predict future memory requests~\cite{Smith,Baer,Stride}.  Modern prefetching
mechanisms are more sophisticated as they look into past memory
behavior~\cite{Address_Correlated,AMPM},
locality~\cite{Spatial_Pattern,SMS,Temporal_Instruction_Fetch,Off_Chip,STMS,SMS_JILP},
control-flow speculation~\cite{BFetch,MTBFetch}, and many other aspects to
detect complex access patterns in modern applications.

@@

%% Simple stride prefetching techniques only detect sequences of addresses that differ by a constant value and fail to capture diverse memory access patterns. 
Lookahead prefetchers can learn complicated patterns by collecting histories of
observed data access patterns.
%EB: Didn't really understand this: 
%% "and correlating these with the strides, by considering each new page in a vacuum."
Kim {\em et. al.} proposed
SPP~\cite{SPP}, a confidence-based lookahead prefetcher. SPP creates a
signature associated to a memory request by compressing the history of
accesses. By correlating the signature with future likely delta patterns, SPP
learn both simple and complicated memory access patterns quickly. Our proposed
technique builds upon SPP. Thus, we describe SPP in more detail.

### 2018-12-05 related.tex
BG and Related

Perceptron-based prediction mechanisms have proven to yield superior accuracy, which represents an opportunity for optimizations in cache management. 
Khan \textit{et. al.} proposed SDBP \cite{sdbp}, in this work, a sampler structure keeps partial tags of sampled sets separate from the cache. Three tables of two-bit saturating counters are accessed using a technique similar to a skewed branch predictor \cite{Piece_Linear}. For each hit to a block in the sampled set, the program counter (PC) of the relevant memory instruction is hashed into the three tables and the corresponding counters are decremented. For each eviction from a sampled set, the counters corresponding to the PC of the last instruction to access the victim block are incremented. For an LLC access, the predictor is consulted by hashing the PC of the memory access instruction into the three tables and taking the sum of the indexed counters. When the sum exceeds some threshold, the accessed block is predicted to be dead. Tags in sampled sets are managed with true LRU and a reduced associativity, but the LLC may be managed by any policy. The paper applies its predictor to replacement and bypass.

Teran \textit{et al.} propose using perceptron learning for reuse prediction driving a bypass and replacement optimization \cite{Perc_Reuse}. In that work, perceptron learning is used to set weights selected by hashes of multiple features consisting of the PC, several recent PCs, and two different shifts of the tag of the referenced block. Each feature is hashed to index its own table of weights. The selected weights are summed and thresholded to make a prediction. When a sampled block is reused or evicted, the corresponding weights are decremented or incremented, respectively, according to the perceptron learning rule.

The continuation of Perceptron was later introduced in the work, Multiperspective Reuse Prediction \cite{Multiperspective}. Compared to previous work that focuses on one or two features at a time, Multiperspective prediction uses many features, each contributing to the overall prediction. The prediction enables placement, promotion and bypass optimizations that improves performance.

@@

Two widely used prefetchers are Next-n Lines \cite{nextnlines} and Stride \cite{stride}, both of which capture regular memory access patterns with very low hardware overhead. The Next-n Lines prefetcher simply queues prefetches for the next n lines after any given miss, under the expectation that the principle of spatial locality will hold and those cache lines after a missed line are likely to be used in the future. The stride prefetcher is slightly more sophisticated; it attempts to identify simple stride reference patterns in programs based upon the past behavior of missing loads. Similar to the Next-n Lines technique, when a given load misses, cache lines ahead of that miss are fetched in the pattern following the previous behavior in the hope of avoiding future misses. Both prefetchers have been widely used due to the simple design and low hardware overhead. However, without further knowledge about temporal locality and application characteristics, these prefetchers cannot do more than detecting and prefetching regular memory access patterns with limited spatial locality.

Somogyi et al. proposed one of the top practical, low-overhead prefetchers, the Spatial Memory Streaming (SMS) prefetcher \cite{sms}. SMS leverages code-based correlation to take advantage of spatial locality in the applications over larger regions of memory (called spatial regions). It predicts the future access pattern within a spatial region around a miss, based on a history of access patterns initiated by that missing instruction in the past. While the SMS prefetcher is effective, it is indirectly inferring future program control-flow when it speculates on the misses in a spatial region. As a result, the state overheads of this predictor can be higher than the others in this class.

\subsection{Lookahead Prefetchers}

To address both prefetching coverage and accuracy, prior work has adopted lookahead 
mechanisms. Lookahead prefetchers efficiently encode the relationship between accesses to yield future predictions, enabling further speculative lookahead accesses.
VLDP \cite {vldp} prefetches a static depth ahead of the demand fetch, without considering the prefetching confidence,

DA-AMPM

While previous work had proposed dynamically finding the best prefetch offset, they have failed to look into prefetch timeliness. Michaud proposed the Best-Offset prefetcher \cite{BOP}. A prefetch offset is set automatically and dynamically tries to adapt to the application behavior. A good prefetch offset is determined by looking at two recent access to a particularly line that do not happen too recently. The time between both accesses should be greater than the latency for completing a prefetch request.

@@

A low-accuracy aggressive prefetcher can significantly harm performance. In order to minimize the interference from prefetching, Wu \textit{et. al} proposed PACMan ~\cite{pacman}, a prefetch-aware cache management policy. PACMan dedicated a few sets of the LLC to each of the three competing policies that treat demand and prefetch request differently and uses the policy that shows the lowest number of cache misses. Similarly, Seshadri  \textit{et. al.} propose ICP \cite{icp}. ICP is also designed as a comprehensive mechanism to mitigate prefetching interference. ICP simply demotes a prefetch blow to the lowest priority on a demand hit based on the observation that most prefetches are dead after their first hit. To address prefetcher-caused cache pollution, it also uses a variation of EAF [27] to track prefetching accuracy and inserts only an accurate prefetch to the higher priority position in the LRU stack. ICP assumes, however, that all prefetches are inserted only into the LLC, which restricts the maximum benefit of prefetching.

>>>>

Perceptron-based prediction mechanisms have proven to yield superior accuracy, 
which represents an opportunity for optimizations in cache management. 
Khan \textit{et. al.} proposed SDBP~\cite{sdbp}, in this work, a sampler 
structure keeps partial tags of sampled sets separate from the cache. 
Three tables of two-bit saturating counters are accessed using a technique 
similar to a skewed branch predictor~\cite{Piece_Linear}. For each hit to 
a block in the sampled set, the program counter (PC) of the relevant memory 
instruction is hashed into the three tables and the corresponding counters 
are decremented. For each eviction from a sampled set, the counters corresponding 
to the PC of the last instruction to access the victim block are incremented. 
For an LLC access, the predictor is consulted by hashing the PC of the memory 
access instruction into the three tables and taking the sum of the indexed 
counters. When the sum exceeds some threshold, the accessed block is 
predicted to be dead. Tags in sampled sets are managed with true LRU and 
a reduced associativity, but the LLC may be managed by any policy. 
The paper applies its predictor to do block replacement and bypass for LLC.

Teran \textit{et al.} propose using perceptron learning for reuse prediction 
driving a bypass and replacement optimization~\cite{Perc_Reuse}. In that work, 
perceptron learning is used to set weights selected by hashes of multiple features 
consisting of the PC, several recent PCs, and two different shifts of the tag 
of the referenced block. Each feature is hashed to index its own table of weights. 
The selected weights are summed and thresholded to make a prediction. When a sampled 
block is reused or evicted, the corresponding weights are decremented or incremented, 
respectively, according to the perceptron learning rule.

The continuation of Perceptron was later introduced in the work, Multiperspective Reuse 
Prediction~\cite{Multiperspective}. Compared to previous work that focuses on one or two 
features at a time, Multiperspective prediction uses many features, each contributing 
to the overall prediction. The prediction enables placement, promotion and bypass 
optimizations that further improves performance.

@@

Two widely used prefetchers are Next-n Lines~\cite{nextn} and Stride~\cite{stride}
prefetchers, both of which capture regular memory access patterns with very low hardware overhead. 
The Next-n Lines prefetcher simply queues prefetches for the next n lines after any given miss, 
under the expectation that the principle of spatial locality will hold and those cache lines after 
a missed line are likely to be used in the future. The stride prefetcher is slightly more 
sophisticated -- it attempts to identify simple stride reference patterns in programs based 
upon the past behavior of missing loads. Similar to the Next-n Lines technique, 
when a given load misses, cache lines ahead of that miss are fetched in the pattern 
following the previous behavior in the hope of avoiding future misses. Both prefetchers 
have been widely used due to the simple design and low hardware overhead. However, 
without further knowledge about temporal locality and application characteristics, 
these prefetchers cannot do more than detecting and prefetching regular memory access 
patterns with limited spatial locality.

Somogyi et al. proposed one of the top practical, low-overhead prefetchers, the Spatial 
Memory Streaming (SMS) prefetcher~\cite{SMS}. SMS leverages code-based correlation 
to take advantage of spatial locality in the applications over larger regions of memory 
(called spatial regions). It predicts the future access pattern within a spatial 
region around a miss, based on a history of access patterns initiated by that 
missing instruction in the past. While the SMS prefetcher is effective, it is 
indirectly inferring future program control-flow when it speculates on the misses 
in a spatial region. As a result, the state overheads of this predictor can 
be higher than the others in this class. This idea was further extended by 
the class of related prefetchers~\cite{STMS,Temporal_Instruction_Fetch}.

While previous work had proposed dynamically finding the best prefetch offset, they have failed to 
look into prefetch timeliness. Michaud proposed the Best-Offset prefetcher~\cite{BOP}. 
A prefetch offset is set automatically and dynamically tries to adapt to the application behavior. 
A good prefetch offset is determined by looking at two recent access to a particularly line 
that did not happen too recently. The time between both accesses should be 
greater than the latency for completing a prefetch request.

\subsection{Lookahead Prefetchers}
While the stride and the next-n perefetchers are efficient, they suffer from their limited
ability to capture memory delta patterns. Moreover, a good prefetch proves to be useful only 
if the prefetcher manages to bring the block before its next demand access. Both these issues 
can be overcome by Lookahead Prefetchers. These prefetchers try to use their own predictions 
to create the spelative path of the program's memory access pattern. By going down the speculation
trail till a certain depth, the prefetcher can bring in cache lines well before they are demand 
accessed by the processor. VLDP~\cite{VLDP} prefetches a static depth ahead of the demand fetch, 
without considering the prefetching confidence. More recent Lookahead Prefetchers include SPP and
KPC~\cite{KPC}

Talk about DA-AMPM

@@

A low-accuracy aggressive prefetcher can significantly harm performance. 
In order to minimize the interference from prefetching, Wu \textit{et. al} proposed PACMan~\cite{pacman}, 
a prefetch-aware cache management policy. PACMan dedicated a few sets of the LLC to each of the three 
competing policies that treat demand and prefetch request differently and uses the policy that 
shows the lowest number of cache misses. Similarly, Seshadri  \textit{et. al.} proposed ICP~\cite{icp}. 
ICP is also designed as a comprehensive mechanism to mitigate prefetching interference. It simply 
demotes a prefetch to the lowest priority on a demand hit based on the observation that most 
prefetches are dead after their first hit. To address prefetcher-caused cache pollution, it also 
uses a variation of EAF~\cite{eaf} to track prefetching accuracy and inserts only an accurate 
prefetch to the higher priority position in the LRU stack. ICP assumes, however, that all 
prefetches are inserted only into the LLC, which restricts the maximum benefit of prefetching.

\subsection{Machine Learning for Prefetching}
Peled \textit{et.al.} introduced
interesting ideas for on-line Reinforcement Learning and dynamically scaling
the magnitude of feedback given to the baseline prefetcher~\cite{Semantics}.
The biggest challenge here is that the prefetcher relies on compiler support for
getting features to build the context.

Liao \textit{et. al.} focused on prefetching for data center
applications~\cite{Datacenter}. They used offline machine learning algorithms
such as SVMs and logistic regression to do a parametric search for an optimal
prefetcher configuration.
